---
title: "p8105_hw3_rh3064"
author: "Rahul Hosalli"
date: '`r Sys.Date()`'
output: github_document
---

```{r include=FALSE}
library(tidyverse)
```

# Problem 2

## Data Loading

```{r}
accel_data <- read_csv("./Data/accel_data.csv")
```

## Data Tidying/Wrangling

The following chunk of code cleans the column names with `janitor::clean_names()`, then using `mutate()` creates a new variable *day_type* which indicates if it was a weekend or weekday. *day* and *day_type* are coerced into factors.

Finally `relocate()` is used to reorder the tibble so that more useful columns are first (*day_id*, *week*, *day*, *day_type*).

`head()` is used to take a glance at the tibble and check the variable classes, which seem reasonable.

```{r}
accel_data <- 
  accel_data %>%
  janitor::clean_names() %>%
  mutate(
    day_type = case_when(day == "Saturday" | day == "Sunday" ~ "Weekend",
                         TRUE ~ "Weekday"),
    day = as.factor(day),
    day_type = as.factor(day_type)
    
    ) %>%
  relocate(day_id, week, day, day_type)
head(accel_data)

accel_long <- accel_data %>% pivot_longer(activity_1:activity_1440,
                                          names_to = "minute",
                                          values_to = "activity_count")

```

## Data Description

There are `r nrow(accel_data)` rows and `r ncol(accel_data)` columns in the *`accel_data`* tibble. There

## Total Activity per Day

```{r}
test<- accel_data %>%
  mutate(
    activity_totalv1 = rowSums(
      across(c(activity_1:activity_1440)))
    )

```

## Plots

```{r}

```

# Problem 3

## Data Loading

```{r}
library(p8105.datasets)
data("ny_noaa")
```

## Data Cleaning

```{r}
noaa_df <- ny_noaa %>% 
  mutate(
    tmax = as.integer(tmax)/10,
    tmin = as.integer(tmin)/10
  ) %>%
  janitor::clean_names() %>%
  separate(date, into = c("year", "month", "dat"), sep = "-")

```
